---
title: "Genotype Imputation: Autoencoder vs IMPUTE2"
author: "Thomas Marshall"
date: 2025-11-24
format:
  html:
    self-contained: true
    fig-width: 6
    fig-asp: 0.618
    toc: true
editor: visual
execute:
  echo: true
  warning: false
---

# Overview

This notebook walks through a small end-to-end pipeline comparing two
approaches to genotype imputation for three SNPs in the *TAS2R38*
bitter-taste receptor gene:

1. **IMPUTE2** — a standard population-genetics imputation tool that
   uses a reference haplotype panel (1000 Genomes) to probabilistically
   infer missing genotypes.
2. **A denoising autoencoder** — a small PyTorch model trained to
   reconstruct a held-out SNP from the dosages of all other SNPs in
   the sample's genotype matrix, learning linkage-disequilibrium
   structure directly from data rather than a reference panel.

The runnable pipeline lives in `R/` and `python/` as standalone
scripts driven by `config.yml`. This notebook reproduces the same
steps narratively, with all output and discussion inline.

All paths below are resolved through `config.yml` — see that file to
point the pipeline at your own data.

# Part 1: IMPUTE2

## 1) Setup

```{r}
library(data.table)
library(tidyverse)
library(yaml)

cfg <- read_yaml(here::here("config.yml"))
base_directory <- cfg$paths$base_directory

impute2_exe <- file.path(base_directory, cfg$paths$impute2_exe)

map_file    <- file.path(base_directory, cfg$paths$ref_panels$map_file)
hap_file    <- file.path(base_directory, cfg$paths$ref_panels$hap_file)
legend_file <- file.path(base_directory, cfg$paths$ref_panels$legend_file)

gens_file   <- file.path(base_directory, cfg$paths$gene_files$gens_file)
sample_file <- file.path(base_directory, cfg$paths$gene_files$sample_file)
strand_file <- file.path(base_directory, cfg$paths$gene_files$strand_file)

output_prefix <- cfg$paths$output_prefix
```

## 2) Running IMPUTE2

```{r}
impute_args <- c(
  "-m", shQuote(map_file),
  "-h", shQuote(hap_file),
  "-l", shQuote(legend_file),
  "-g", shQuote(gens_file),
  "-sample_g", shQuote(sample_file),
  "-strand_g", shQuote(strand_file),
  "-Ne", as.character(cfg$impute2$Ne),
  "-int", as.character(cfg$impute2$interval[1]), as.character(cfg$impute2$interval[2]),
  "-o", shQuote(output_prefix)
)

system2(impute2_exe, args = impute_args)
```

## 3) Inspecting the target variants

```{r}
info_file <- paste0(output_prefix, "_info")
info <- read.table(info_file, header = TRUE, stringsAsFactors = FALSE)

tas_snps <- cfg$target$snps
info_tas <- subset(info, snp_id %in% tas_snps)
info_tas
```

## 4) Extracting dosages

A dosage is the expected allele count at a locus -- a continuous
value between 0 and 2, rather than a hard 0/1/2 genotype call.

```{r}
imputed <- fread(output_prefix, header = FALSE)
imputed_tas <- imputed[V1 %in% tas_snps]

n_samples <- (ncol(imputed_tas) - 5) / 3
n_samples
```

## 5) Mapping sample IDs

```{r}
sample_df <- read.table(sample_file, header = TRUE, stringsAsFactors = FALSE)
sample_ids <- sample_df$ID_2
if (sample_ids[1] == "0") sample_ids <- sample_ids[-1]
```

## 6) Dosage extraction function

```{r}
get_dosage_from_row <- function(row, n_samples) {
  row_vec <- as.numeric(row)
  probs <- row_vec[6:(5 + 3 * n_samples)]
  prob_mat <- matrix(probs, ncol = 3, byrow = TRUE)
  # dosage = 0 * P(AA) + 1 * P(AB) + 2 * P(BB)
  as.numeric(prob_mat %*% c(0, 1, 2))
}
```

## 7) Dosages for the three TAS2R38 variants

```{r}
dosage_list <- lapply(tas_snps, function(snp) {
  get_dosage_from_row(imputed_tas[V1 == snp], n_samples)
})

tas_dosages <- data.frame(sample_id = sample_ids)
for (i in seq_along(tas_snps)) {
  tas_dosages[[tas_snps[i]]] <- dosage_list[[i]]
}

head(tas_dosages)

write.csv(tas_dosages, cfg$paths$impute2_dosages_csv, row.names = FALSE)
```

# Part 2: Autoencoder

## 1) Building the full genotype matrix

```{r}
imputed <- fread(output_prefix, header = FALSE)

dos_mat <- t(apply(imputed, 1, get_dosage_from_row, n_samples = n_samples))
colnames(dos_mat) <- sample_ids
rownames(dos_mat) <- imputed$V1

X <- t(dos_mat)  # rows = samples, cols = SNPs

write.csv(X, file = cfg$paths$genotype_matrix_csv, row.names = TRUE)
```

## 2) Training the autoencoder

The model is a small encoder-decoder network. During training, the
target SNP (`exm663431`) is randomly hidden for a fraction of samples
in each batch, and loss is computed only at those hidden positions --
this is what forces the network to learn to predict it from the
surrounding genotype context rather than memorizing it.

The full implementation lives in `python/autoencoder.py` and
`python/train.py`; the cell below mirrors that logic for illustration.

```{python}
import sys
sys.path.append("python")

import yaml
import numpy as np
import pandas as pd
import torch

from autoencoder import (
    AEConfig, normalize, build_missing_mask,
    train_autoencoder, impute_target_snp
)

with open("config.yml") as f:
    cfg = yaml.safe_load(f)

df = pd.read_csv(cfg["paths"]["genotype_matrix_csv"], index_col=0)
sample_ids = df.index.to_list()
snp_ids = df.columns.to_list()

X_np = df.to_numpy().astype(np.float32)
X_norm, snp_mean, snp_std = normalize(X_np)
missing_mask = build_missing_mask(X_norm)
X_norm_filled = np.nan_to_num(X_norm, nan=0.0)

target_snp = cfg["target"]["hide_snp"]
target_j = snp_ids.index(target_snp)

ae_cfg = AEConfig(**cfg["autoencoder"])
device = torch.device("cuda" if torch.cuda.is_available() else "cpu")

model = train_autoencoder(X_norm_filled, missing_mask, target_j, ae_cfg, device)
```

## 3) Generating imputed dosages

```{python}
recon_dosages = impute_target_snp(
    model, X_norm_filled, target_j, snp_mean, snp_std, device
)

ae_full_df = pd.DataFrame(recon_dosages, index=sample_ids, columns=snp_ids)

tas_snps_py = cfg["target"]["snps"]
ae_dos_df = ae_full_df[tas_snps_py].copy()
ae_dos_df.insert(0, "sample_id", sample_ids)

ae_dos_df.to_csv(cfg["paths"]["ae_dosages_csv"], index=False)
ae_dos_df.head()
```

# Part 3: Autoencoder vs IMPUTE2

```{r}
library(ggplot2)

tas_imp <- fread(cfg$paths$impute2_dosages_csv)
ae_raw  <- fread(cfg$paths$ae_dosages_csv)

ae_tas <- ae_raw[, c("sample_id", tas_snps), with = FALSE]

common_ids <- intersect(tas_imp$sample_id, ae_tas$sample_id)
tas_imp_f <- tas_imp[sample_id %in% common_ids]
ae_tas_f  <- ae_tas[sample_id %in% common_ids]

setkey(tas_imp_f, sample_id)
setkey(ae_tas_f,  sample_id)
tas_imp_f <- tas_imp_f[order(sample_id)]
ae_tas_f  <- ae_tas_f[order(sample_id)]

stopifnot(identical(tas_imp_f$sample_id, ae_tas_f$sample_id))

imp_long <- melt(tas_imp_f, id.vars = "sample_id", variable.name = "snp", value.name = "dos_imp")
ae_long  <- melt(ae_tas_f,  id.vars = "sample_id", variable.name = "snp", value.name = "dos_ae")

comp <- merge(imp_long, ae_long, by = c("sample_id", "snp"))
comp[, diff := dos_ae - dos_imp]

rmse_overall <- sqrt(mean(comp$diff^2, na.rm = TRUE))
cor_overall  <- cor(comp$dos_imp, comp$dos_ae, use = "complete.obs")

rmse_overall
cor_overall

comp_stats <- comp[, .(
  cor  = cor(dos_imp, dos_ae, use = "complete.obs"),
  rmse = sqrt(mean((dos_ae - dos_imp)^2, na.rm = TRUE))
), by = snp]

comp_stats
```

## Genotype accuracy (hard calls)

```{r}
to_geno <- function(d) {
  ifelse(d < 0.5, 0L, ifelse(d < 1.5, 1L, 2L))
}

comp[, geno_imp := to_geno(dos_imp)]
comp[, geno_ae  := to_geno(dos_ae)]

acc_overall <- mean(comp$geno_imp == comp$geno_ae, na.rm = TRUE)
acc_overall

acc_by_snp <- comp[, .(accuracy = mean(geno_imp == geno_ae, na.rm = TRUE)), by = snp]
acc_by_snp
```

## Diagnostic plots

```{r}
ggplot(comp[snp == "exm663431"], aes(x = dos_imp, y = dos_ae)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "IMPUTE2 vs Autoencoder Dosages for exm663431",
    x = "IMPUTE2 dosage",
    y = "Autoencoder dosage"
  )

ggplot(comp, aes(x = (dos_imp + dos_ae) / 2, y = diff, color = snp)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  labs(
    title = "Difference: Autoencoder - IMPUTE2 Dosages",
    x = "Mean dosage (IMPUTE2 & AE)",
    y = "Difference (AE - IMPUTE2)"
  )

sub <- comp[snp %in% tas_snps]
sub[, err_ae := abs(dos_imp - dos_ae)]

ggplot(sub, aes(x = snp, y = err_ae)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Imputation Error for exm663431",
    subtitle = "Autoencoder vs IMPUTE2",
    x = "SNP identity",
    y = "Absolute error (autoencoder)"
  ) +
  theme_minimal()
```

# Discussion

The autoencoder's accuracy depends heavily on how much linkage
disequilibrium exists between the target SNP and the rest of the
genotyped variants in the window -- it has no access to the 1000
Genomes reference panel that gives IMPUTE2 its statistical power, so
agreement between the two should be read as a rough concordance check
rather than a ground-truth validation. A natural next step is testing
this approach on a larger SNP window and a larger sample set, where
the autoencoder has more LD signal to exploit.
